---
title: "Post-processing cTWAS results"
author: "Kaixuan Luo, Sheng Qian"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-processing cTWAS results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.width = 6,
                      fig.height = 4,
                      fig.align = "center",
                      fig.cap = "&nbsp;",
                      dpi = 120)
```


A few issues can potentially lead to problematic results by cTWAS. In this tutorial, we will show how to address these issues by performing post-processing of cTWAS results.

Load the packages. 
```{r load_packages, message=FALSE}
library(ctwas)
```


Let's first load the cTWAS input data needed to run this tutorial: 
GWAS sample size (`gwas_n`), preprocessed GWAS z-scores (`z_snp`), prediction models of molecular traits (`weights`), and the reference data: including information about regions (`region_info`), reference variant list (`snp_map`). 

```{r load_ctwas_input}
gwas_n <- 343621

z_snp <- readRDS(system.file("extdata/sample_data", "LDL_example.preprocessed.z_snp.RDS", package = "ctwas"))

weights <- readRDS(system.file("extdata/sample_data", "LDL_example.preprocessed.weights.RDS", package = "ctwas"))

region_info <- readRDS(system.file("extdata/sample_data", "LDL_example.region_info.RDS", package = "ctwas"))

snp_map <- readRDS(system.file("extdata/sample_data", "LDL_example.snp_map.RDS", package = "ctwas"))
```


When running with LD, we load the cTWAS result from running the `ctwas_sumstats()` function. We also load reference LD (`LD_map`) and the directory where we save the correlation matrices.

```{r load_ctwas_res}
ctwas_res <- readRDS(system.file("extdata/sample_data", "LDL_example.ctwas_sumstats_res.RDS", package = "ctwas"))
z_gene <- ctwas_res$z_gene
param <- ctwas_res$param
finemap_res <- ctwas_res$finemap_res
region_data <- ctwas_res$region_data
screen_res <- ctwas_res$screen_res

LD_map <- readRDS(system.file("extdata/sample_data", "LDL_example.LD_map.RDS", package = "ctwas"))

cor_dir <- system.file("extdata/sample_data", "cor_matrix", package = "ctwas")
```


When running without LD, we load the cTWAS result from running the `ctwas_sumstats_noLD()` function:

```{r load_ctwas_noLD_res, eval=FALSE}
ctwas_res <- readRDS(system.file("extdata/sample_data", "LDL_example.ctwas_sumstats_noLD_res.RDS", package = "ctwas"))
z_gene <- ctwas_res$z_gene
param <- ctwas_res$param
finemap_res <- ctwas_res$finemap_res
region_data <- ctwas_res$region_data
screen_res <- ctwas_res$screen_res
```


##  Cross-region LD

If the variants in a prediction model of a gene (or molecular trait) span two (or more) regions, it would be unclear to cTWAS what region the gene should be assigned to. If this happens, cTWAS will attempt to assign the gene to one of the two regions that contain most of the weights. Nevertheless, there will be a risk of cross-region LD, where the genetic component of this gene may correlate with variants or genes of both regions. This violates the assumption of cTWAS and can lead to false positive findings. 

We address this “cross-region” problem by performing “region merging” as a post-processing procedure. If any gene has variants in the weights that span two or more regions ("cross-boundary"), those regions will be merged, and cTWAS will rerun the fine-mapping step on the merged regions. 

This post-processing procedure contains multiple steps, as described in the section "Post-processing steps for region merging" below. To simplify these steps, we have the functions `postprocess_region_merging()` and `postprocess_region_merging_noLD()`, which automate the region merging steps with or without LD.

Region merging with LD: 

```{r postprocess_region_merging, eval=FALSE}
postprocess_region_merging_res <- postprocess_region_merging(
                                      region_info,
                                      region_data,
                                      z_snp,
                                      z_gene,
                                      weights,
                                      LD_map,
                                      snp_map,
                                      finemap_res = finemap_res, 
                                      susie_alpha_res = susie_alpha_res,
                                      mapping_table = mapping_table,
                                      L = 5,
                                      group_prior = group_prior,
                                      group_prior_var = group_prior_var,
                                      pip_thresh = 0.5,
                                      filter_cs = FALSE,
                                      maxSNP = 20000,
                                      save_cor = TRUE,
                                      cor_dir = "./cor_matrix"
                                      ncore = 6)
```


Region merging without LD: 


```{r postprocess_region_merging_noLD, eval=FALSE}
postprocess_region_merging_noLD_res <- postprocess_region_merging_noLD(
                                           region_info,
                                           region_data,
                                           z_snp,
                                           z_gene,
                                           weights,
                                           snp_map,
                                           finemap_res = finemap_res, 
                                           susie_alpha_res = susie_alpha_res,
                                           mapping_table = mapping_table,
                                           group_prior = group_prior,
                                           group_prior_var = group_prior_var,
                                           pip_thresh = 0.5,
                                           filter_cs = FALSE,
                                           maxSNP = 20000,
                                           ncore = 6)
```


If you would like to run each of the region merging steps separately, please see the section below.

### Post-processing steps for region merging

**Step 1. Select boundary genes**

We first select the boundary genes to merge. This could be all genes whose weights span two or more regions. In practice, the genes that are unlikely to be causal for the phenotype generally wouldn’t cause problems. So we can limit to the cross-boundary genes that are plausible risk genes. We can use the following function to get boundary genes and select those above a PIP cutoff (default = 0.5). 

```{r select_boundary_genes, eval=FALSE}
res <- select_boundary_genes(region_info, 
                             weights, 
                             gene_ids = z_gene$id, 
                             finemap_res = finemap_res, 
                             susie_alpha_res = susie_alpha_res,
                             mapping_table = mapping_table,
                             pip_thresh = 0.5,
                             filter_cs = FALSE)
boundary_genes <- res$boundary_genes
selected_boundary_genes <- res$selected_boundary_genes
```


We can also limit to those in credible sets by setting `filter_cs = TRUE`. 

**Step 2. Merge regions**

Next, we use the function `merge_region_data()` to perform region merging.  It first identifies overlapping regions from the selected genes, then creates the data about the merged region(s) needed for fine-mapping. The data includes:  `merged_region_data`,  `merged_region_info`, and the variant and LD information of the merged region(s), `merged_snp_map` and `merged_LD_map`. The function expands the merged regions with all SNPs if the `region_data` contains thinned SNPs. It also creates `merged_region_id_map`, a data frame containing region IDs for merged regions and the original region IDs, so that one could keep track of the regions that were merged.  

```{r merge_region_data, eval=FALSE}
merge_region_res <- merge_region_data(selected_boundary_genes,
                                       region_data,
                                       region_info = region_info,
                                       LD_map = LD_map,
                                       snp_map = snp_map,
                                       weights = weights,
                                       z_snp = z_snp,
                                       z_gene = z_gene,
                                       maxSNP = 20000)
merged_region_data <- merge_region_res$merged_region_data
merged_region_info <- merge_region_res$merged_region_info
merged_LD_map <- merge_region_res$merged_LD_map
merged_snp_map <- merge_region_res$merged_snp_map
merged_region_id_map <- merge_region_res$merged_region_id_map
```


We could set `maxSNP` to limit the number of SNPs in the region, if merged regions contain too many SNPs. 

When running without LD, we could merge regions using the function `merge_region_data_noLD()`. 

```{r merge_region_data_noLD, eval=FALSE}
merge_region_res <- merge_region_data_noLD(selected_boundary_genes,
                                           region_data,
                                           region_info = region_info,
                                           snp_map = snp_map,
                                           z_snp = z_snp,
                                           z_gene = z_gene,
                                           maxSNP = 20000)
merged_region_data <- merge_region_res$merged_region_data
merged_region_info <- merge_region_res$merged_region_info
merged_snp_map <- merge_region_res$merged_snp_map
merged_region_id_map <- merge_region_res$merged_region_id_map
```



**Step 3. Fine-map merged regions**

Next, we run fine-mapping again for these merged regions, similar to the section "Fine-mapping screened regions" in the tutorial ["Running cTWAS analysis"][running_ctwas_analysis_tutorial].

```{r finemap_merged_regions, eval=FALSE}
res <- finemap_regions(merged_region_data,
                       LD_map = merged_LD_map,
                       weights = weights,
                       group_prior = group_prior,
                       group_prior_var = group_prior_var,
                       L = 5,
                       save_cor = TRUE,
                       cor_dir = "./cor_matrix")
merged_region_finemap_res <- res$finemap_res
merged_region_susie_alpha_res <- res$susie_alpha_res
```


The output of this function has the same format as those from regular cTWAS analysis. It returns fine-mapping results for merged regions. 

We could also run fine-mapping without LD on the merged region data:

```{r screen_merged_regions_noLD, eval=FALSE}
res <- finemap_regions_noLD(merged_region_data,                                                     
                            group_prior = group_prior,
                            group_prior_var = group_prior_var)
merged_region_finemap_res <- res$finemap_res
merged_region_susie_alpha_res <- res$susie_alpha_res
```


**Step 4. Update fine-mapping results after region merging**

We can update the fine-mapping results after region merging:

```{r update_merged_region_finemap_res, eval=FALSE}
res <- update_merged_region_finemap_res(finemap_res, 
                                        susie_alpha_res, 
                                        merged_region_finemap_res, 
                                        merged_region_susie_alpha_res,
                                        merged_region_id_map)
updated_finemap_res <- res$finemap_res
updated_susie_alpha_res <- res$susie_alpha_res
```


After region merging, we can also update the region data (and other input data), which could be used in later, e.g. LD mismatch analysis:

```{r update_merged_region_data, eval=FALSE}
updated_data_res <- update_merged_region_data(region_data, merged_region_data,
                                              region_info, merged_region_info,
                                              LD_map, merged_LD_map,
                                              snp_map, merged_snp_map,
                                              merged_region_id_map)
updated_region_data <- updated_data_res$updated_region_data
updated_region_info <- updated_data_res$updated_region_info
updated_LD_map <- updated_data_res$updated_LD_map
updated_snp_map <- updated_data_res$updated_snp_map
```


Similarly, we could update region data when running without LD:

```{r update_merged_region_data_noLD, eval=FALSE}
updated_data_res <- update_merged_region_data_noLD(region_data, merged_region_data,
                                                   region_info, merged_region_info,
                                                   snp_map, merged_snp_map,
                                                   merged_region_id_map)
updated_region_data <- updated_data_res$updated_region_data
updated_region_info <- updated_data_res$updated_region_info
updated_snp_map <- updated_data_res$updated_snp_map
```


## LD mismatch analysis

LD mismatch between GWAS data (in-sample LD) and the reference LD could lead to false positives in fine-mapping. Diagnostic tools including [SuSiE-RSS][susierss_diagnostic], and [DENTIST][DENTIST], have been developed to check possible LD mismatch. Because it is very time consuming to run the LD mismatch diagnosis for all the regions across the genome, we will perform LD mismatch diagnosis and adjustment only for selected regions with high PIP signals in the post-processing. 

Here, we perform the LD mismatch diagnosis using [SuSiE-RSS][susierss_diagnostic] for selected regions. It is an optional step that users could run after finishing the main cTWAS analysis. 

Similar to region merging, the post-processing procedure for LD mismatch analysis also involves several steps. We have a function `postprocess_LD_mismatch()`, which automates the post-processing steps for LD mismatch analysis.

```{r postprocess_LD_mismatch, eval=FALSE}
postprocess_LD_mismatch_res <- postprocess_LD_mismatch(
                                   region_data,
                                   z_snp,
                                   weights,
                                   LD_map,
                                   snp_map,
                                   gwas_n,
                                   finemap_res,
                                   susie_alpha_res,
                                   group_prior = group_prior,
                                   group_prior_var = group_prior_var,
                                   min_nonSNP_PIP = 0.8,
                                   filter_cs = FALSE,
                                   p_diff_thresh = 5e-6,
                                   pip_thresh = 0.5,
                                   plot = TRUE,
                                   maxSNP = 20000,
                                   ncore = 6)
```


Please see the details of LD mismatch analysis below if you would like to perform each step separately.

*Note*: In practice, it could be helpful to first run the region merging procedure (with `postprocess_region_merging()`), and then perform the LD mismatch analysis (with `postprocess_LD_mismatch()`) using the outputs from region merging. 


### Post-processing steps for LD mismatch analysis

**Step 1. Select regions of interest**

Users first need to choose regions of interest to run LD-mismatch analysis. For example, we could use the function `compute_region_nonSNP_PIPs()` to compute total non-SNP PIPs for the regions. The non-SNP PIP of a region is the sum of PIPs of all molecular traits in that region. We can then select regions with total non-SNP PIPs > 0.8 to run LD mismatch diagnosis:

```{r select_regions_to_rerun, eval=FALSE}
nonSNP_PIPs <- compute_region_nonSNP_PIPs(finemap_res, filter_cs = FALSE)
selected_region_ids <- names(nonSNP_PIPs)[nonSNP_PIPs > 0.8]
```


We could also limit to credible sets by setting  `filter_cs = TRUE`. 

**Step 2. Diagnose LD mismatch**

We use the function `diagnose_LD_mismatch_susie()` to perform LD mismatch diagnosis for these regions.  This function uses [SuSiE-RSS][susierss_diagnostic] to detect problematic SNPs among all variants in a region. Basically, it infers the expected statistic of a variant, based on the statistics of nearby variants and their LD from the reference, and compares this with the observed statistic. Inconsistency between the two would suggest potential LD mismatch.  

```{r diagnose_LD_mismatch, eval=FALSE}
res <- diagnose_LD_mismatch_susie(region_ids = selected_region_ids, 
                                  z_snp = z_snp, 
                                  LD_map = LD_map, 
                                  gwas_n = gwas_n,
                                  plot = TRUE,
                                  p_diff_thresh = 5e-6,
                                  ncore = 4)
problematic_snps <- res$problematic_snps
flipped_snps <- res$flipped_snps
condz_stats <- res$condz_stats
plots <- res$plots
```


This returns a list of problematic SNPs (diagnostic test `p-value < 5e-6`, by default), flipped SNPs, and the test statistics of `kriging_rss` function from [SuSiE-RSS][susierss_diagnostic]. When `plot = TRUE`, it returns the plots with expected vs. observed z-scores for diagnosis. 

**Step 3. Fine-map without LD for regions with problematic genes**

Our basic strategy of dealing with LD mismatch is: we use the list of problematic SNPs to identify the molecular traits whose results may be affected by LD-mismatch. We would then run fine-mapping without LD (L = 1) in the regions containing these molecular traits, assuming a single causal signal in such a region. The fine-mapping results in this setting would be independent of LD. 

We choose the genes with some plausibility of being risk genes (gene PIP > 0.5 by default) and problematic SNPs in their weights. We would then select regions containing these problematic genes.

```{r problematic_region_ids, eval=FALSE}
problematic_genes <- get_problematic_genes(problematic_snps, 
                                           weights, 
                                           finemap_res, 
                                           pip_thresh = 0.5)
```

We then rerun the fine-mapping without LD information for the problematic regions, using the  `finemap_regions_noLD()` function:
 
```{r rerun_finemap_problematic_regions, eval=FALSE}
if (length(problematic_genes) > 0) {
  problematic_region_ids <- unique(finemap_res$region_id[finemap_res$id %in% problematic_genes])
  rerun_region_data <- screen_res$screened_region_data[problematic_region_ids]
  res <- finemap_regions_noLD(rerun_region_data, 
                              group_prior = group_prior,
                              group_prior_var = group_prior_var)
  rerun_finemap_res <- res$finemap_res
  rerun_susie_alpha_res <- res$susie_alpha_res
}
```


We can check the fine-mapping results (without LD) for the problematic genes:
```{r check_finemap_rerun_res, eval=FALSE}
rerun_finemap_res[rerun_finemap_res$id %in% problematic_genes,]
```


We can update the fine-mapping results for the problematic regions:
```{r update_finemap_noLD_res, eval=FALSE}
res <- update_finemap_res(finemap_res, 
                          susie_alpha_res, 
                          rerun_finemap_res, 
                          rerun_susie_alpha_res,
                          updated_region_ids = problematic_region_ids)
updated_finemap_res <- res$finemap_res
updated_susie_alpha_res <- res$susie_alpha_res
```



[reference]: https://xinhe-lab.github.io/ctwas/reference/index.html
[UKBB_LD_ref]: https://uchicago.box.com/s/jqocacd2fulskmhoqnasrknbt59x3xkn
[LDetect]: https://github.com/endrebak/ldetect
[PredictDB]: http://predictdb.org/
[FUSION_format]: http://gusevlab.org/projects/fusion/#compute-your-own-predictive-models
[S-PrediXcan]: https://www.nature.com/articles/s41467-018-03621-1
[susierss_diagnostic]: https://stephenslab.github.io/susieR/articles/susierss_diagnostic.html
[DENTIST]: https://github.com/Yves-CHEN/DENTIST/
[minimal_tutorial]: https://xinhe-lab.github.io/multigroup_ctwas/articles/minimal_tutorial.html
[preparing_input_data_tutorial]: https://xinhe-lab.github.io/multigroup_ctwas/articles/preparing_input_data.html
[running_ctwas_analysis_tutorial]: https://xinhe-lab.github.io/multigroup_ctwas/articles/running_ctwas_analysis.html
[summarizing_results_tutorial]: https://xinhe-lab.github.io/multigroup_ctwas/articles/summarizing_results.html
[postprocessing_tutorial]: https://xinhe-lab.github.io/multigroup_ctwas/articles/postprocessing.html
[utility_functions_tutorial]:https://xinhe-lab.github.io/multigroup_ctwas/articles/utility_functions.html


