---
title: "Preparing cTWAS input data"
author: "Kaixuan Luo, Sheng Qian"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing cTWAS input data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.width = 6,
                      fig.height = 4,
                      fig.align = "center",
                      fig.cap = "&nbsp;",
                      dpi = 120)
```



This tutorial shows how to prepare for the input data for running cTWAS with summary statistics.

Load the package
```{r load_package}
library(ctwas)
```


We use `VariantAnnotation` and `gwasvcf` packages to read the VCF files, so please install those packages before running this tutorial. 

The inputs include GWAS summary statistics, prediction models (referred to as “weights” in the documents), and LD reference. Additionally, cTWAS partitions the genome into disjoint regions, assuming no LD across regions. So a user also needs to provide the definition of these regions.  

Note: In this tutorial, we assume we run full cTWAS using the LD reference. To run cTWAS without LD, see the "minimal tutorial of running cTWAS without LD":

In the tutorial, we use sample data provided in the package.

## GWAS Z-scores

We will use summary statistics from a GWAS of LDL cholesterol in the UK Biobank.  

We can download the VCF from the IEU Open GWAS Project. 
```{bash, eval=FALSE}
wget https://gwas.mrcieu.ac.uk/files/ukb-d-30780_irnt/ukb-d-30780_irnt.vcf.gz
```


cTWAS needs a data frame `z_snp` as input, with columns "id", "A1", "A2", "z". Each row has data of a variant.  We use rsIDs as the variant IDs. The naming of variant IDs should be consistent between summary statistics, LD reference and weights. `A1` is the alternate allele, and `A2` is the reference allele. `z` is the Z-scores of variants in GWAS. In the GWAS summary statistics VCF file we downloaded here, Z-scores are not available, so we need to compute them from the effect sizes and standard errors. 

Let us read the GWAS summary statistics and convert it to the `z_snp` data frame. We will also collect the sample size, which will be used later. 
 
```{r read_sumstats, eval=FALSE}
z_snp <- VariantAnnotation::readVcf("ukb-d-30780_irnt.vcf.gz")
gwas_name <- "ukb-d-30780_irnt"

z_snp <- as.data.frame(gwasvcf::vcf_to_tibble(z_snp))
z_snp$Z <- z_snp$ES/z_snp$SE

gwas_n <- as.numeric(names(sort(table(z_snp$SS),decreasing=TRUE)[1]))
cat("gwas_n=", gwas_n, "\n")


z_snp <- z_snp[,c("rsid", "ALT", "REF", "Z")]
colnames(z_snp) <- c("id", "A1", "A2", "z")
```


If your GWAS summary statistics is of other formats, please extract relevant columns (rsid, alternative allele, reference allele, Z-score) and convert it to the format shown above. 

## Prediction models

Prediction models can be specified in either PredictDB or FUSION format. 
PredictDB is the recommended format, as it has information about correlations among variants included in a model, and also has extra information about the genes in the model. In this user guide, we focus on the PredictDB format. For FUSION format, please see section "FUSION weights" in the tutorial "Utility functions" for details. We also note that it is possible to run cTWAS with only a list of QTLs, which is often the case in practice. See the section "Creating prediction models from QTL lists" in the tutorial "Utility functions"  for details.

In terms of the choice of prediction models, cTWAS performs best when prediction models are sparse, i.e. they have relatively few variants per gene. Dense weights may lead to a “cross-region” problem. Basically, if the variants in the prediction model of a gene span two regions, it would be unclear to cTWAS what region the gene should be assigned to. The problem becomes worse with dense weights. If this happens, cTWAS will attempt to assign the gene to one of the two regions that contain most of the weights. Nevertheless, there will be a risk of cross-region LD which can lead to problems. Given this consideration, we recommend choosing sparse prediction models such as Lasso. If using dense prediction models, we recommend removing variants with weights below a threshold from the prediction models.

Note: we implemented an option to address the “cross-region” problem. It performs “region merging” as a post-processing step. If any gene has variants in the weights that span two regions, those regions will be merged, and cTWAS will rerun the analysis using the merged regions. See the tutorial "Post-processing cTWAS results" for details.   

In this example, we will use liver gene expression models trained on GTEx v8 in the PredictDB format. We can download both the prediction models (.db) and the covariances between variants in the prediction models (.txt.gz) from the link below. The covariances can optionally be used for computing LD.
```{bash, eval=FALSE}
wget https://zenodo.org/record/3518299/files/mashr_eqtl.tar
tar -xvf mashr_eqtl.tar
```


Please check [PredictDB](http://predictdb.org/) for the format of PredictDB weights. To specify weights in PredictDB format, provide the path to the `.db` file when running cTWAS. 

```{r predictdb_weights}
weight_file <- system.file("extdata/sample_data", "expression_Liver.db", package = "ctwas")
```

## Reference data

The reference data include definitions of genomic regions, and the LD reference. Note that the choice of LD reference population is important for fine-mapping. Best practice for fine-mapping is to use an in-sample LD reference (LD computed using the subjects in the GWAS sample). If this is not available, the LD reference should be as representative of the population in the GWAS sample as possible.

It is critical that the genome build (e.g. hg38) of the LD reference matches the genome build used to train the prediction models. The genome build of the GWAS summary statistics does not matter because variant positions are determined by the LD reference.

### Defining regions

cTWAS assumes that the genome is partitioned into approximately independent LD regions. 

We included in the package predefined regions based on European (EUR), Asian (ASN), or African (AFR) populations, using either genome build b38 or b37. 
These regions were previously generated using [LDetect](https://github.com/endrebak/ldetect). 

Here we use the b38 European LDetect blocks, which is again included in the R package.

```{r region_met}
region_file <- system.file("extdata/ldetect", "EUR.b38.ldetect.regions.RDS", package = "ctwas")
region_info <- readRDS(region_file)
region_info <- subset(region_info, chrom == 16)

head(region_info)
```


`region_info` contains the region definitions, and IDs of the regions (by default, we use the convention <chrom_start_stop> for the region IDs).

In this tutorial, we use chr16 as an example, so we specify `chrom = 16`. In real cTWAS analysis, you should run the entire genome. 

### Reference variants and LD matrices

cTWAS performs its analysis region-by-region. The preferred way to run cTWAS is to provide pre-computed LD matrices for each region. If you run cTWAS without using LD matrices, please jump to the section “Mapping reference SNPs to regions” below. 

By default, we need a LD matrix file for each region (preferably in `.RDS` format) and a variant information table accompanying each LD matrix. The `.RDS` files store the LD correlation matrix  (R) for each region (a $p \times p$ matrix, $p$ is the number of variants in the region). The accompanying `.Rvar` files include variant information for the region, and the order of its rows should match the order of rows and columns in the `.RDS` file.

We have precomputed reference LD matrices (`.RDS` files) and the variant information tables (`.Rvar` files) accompanying the LD matrices for UK Biobank. The complete LD matrices of European individuals from UK Biobank can be downloaded [here](https://uchicago.box.com/s/jqocacd2fulskmhoqnasrknbt59x3xkn). 
On the University of Chicago RCC cluster, the b38 reference is available at `/project2/mstephens/wcrouse/UKB_LDR_0.1/` and the b37 reference is available at `/project2/mstephens/wcrouse/UKB_LDR_0.1_b37/`.

The columns of the `.Rvar` file include information on chromosome, variant name, position in base pairs, and the alternative and reference alleles. The `variance` column is the variance of each variant prior to standardization; this is required for PredictDB weights but not FUSION weights. We've also included information on allele frequency in the variant info, but this is optional.

Each variant should be uniquely assigned to a region, and the regions should be left closed and right open, i.e. [start, stop). The positions of the LD matrices must match exactly the positions specified by the region file. Do not include multiallelic variants in the LD reference. 

### Mapping reference SNPs and LD matrices to regions

We will need to link the reference SNP and LD files to the regions. To do this, we add two additional columns "LD_file" and "SNP_file" to `region_info`:  "LD_file" stores the filenames to the precomputed LD (R) matrices (`.RDS` files in our precomputed reference LD files),  "SNP_file" stores the filenames of corresponding variant information (`.Rvar` files). The result is stored in a data structure `region_metatable`. 

The following paths and filenames were from the University of Chicago RCC cluster, please replace them with your own paths. 
```{r add_LD_paths, eval=FALSE}
LD_dir <- "/project2/mstephens/wcrouse/UKB_LDR_0.1/"
genome_version <- "b38"

LD_filestem <- sprintf("ukb_%s_0.1_chr%s.R_snp.%s_%s", genome_version, region_info$chrom, region_info$start, region_info$stop)

region_metatable <- region_info
region_metatable$LD_file <- file.path(LD_dir, paste0(LD_filestem, ".RDS"))
region_metatable$SNP_file <- file.path(LD_dir, paste0(LD_filestem, ".Rvar"))
```


The `create_snp_LD_map()` function takes `region_metatable` as input, extracts the region definitions, ensures that it matches the format required by cTWAS (e.g. adding `region_id` column if not available), and returns updated `region_info`. It reads the variant information from the `SNP_file` column of `region_metatable`, and returns the result as `snp_map`.  It also checks the `LD_file` column of `region_metatable` to make sure the files are available, and returns `LD_map`, a data frame with region IDs and the filenames to the precomputed LD (R) matrices. 

```{r create_snp_LD_map, eval=FALSE}
res <- create_snp_LD_map(region_metatable)
region_info <- res$region_info
snp_map <- res$snp_map
LD_map <- res$LD_map
```



### Mapping reference SNPs to regions

When preparing data without LD, we need `region_info` (region definitions, no need to provide "LD_file" and "SNP_file" columns) and `ref_snp_info`, a data frame containing information of all variants from the reference. 

We have the reference variant information from UK Biobank European population in [hg38](https://uchicago.box.com/s/t089or92dkovv0epkrjvxq8r9db9ys99) and [hg19](https://uchicago.box.com/s/ufko2gjagcb693dob4khccqubuztb9pz).

We then use the `create_snp_map()` function to preprocess region definitions, and map variants from the LD reference to regions, which returns processed `region_info` and `snp_map`.

```{r create_snp_map, eval=FALSE}
ref_snp_info_file <- system.file("extdata/sample_data", "ukb_b38_0.1_chr16_var_info.Rvar.gz", package = "ctwas")

ref_snp_info <- fread(ref_snp_info_file, sep = "\t")
class(ref_snp_info) <- "data.frame"

res <- create_snp_map(region_info, ref_snp_info)
region_info <- res$region_info
snp_map <- res$snp_map
```



## Data harmonization

There are a few potential problems when preparing the input data. First, the variants in the three sets of input data, GWAS summary statistics, weights, and the reference data, may not match. Only variants in all three input data will be used in cTWAS. The `preprocess_weights()` function described below will automatically select variants in all three input data. So it is important to maximize the overlap of the variants in the three sets. This can be done for example, by imputing GWAS summary statistics of the variants missing in GWAS but in the LD reference. Another useful pre-processing step is to perform Minor allele frequency (MAF) filtering on the GWAS data so that only those with MAF above a certain cutoff would be used in the analysis, ideally the same cutoff used in the references. Additionally, when building the prediction models of gene expression, it is better to impute the genotype data using the LD reference, if possible.  All these steps should be done before running cTWAS.  

The second potential problem is that the effect alleles in the prediction model, GWAS and LD reference may not agree with each other, thus we need to ``harmonize'' the data to ensure that the effect alleles match in all three data sources. 
    
Another potential problem is the LD of the GWAS data (in-sample LD) do not match the reference LD. This can lead to problems in fine-mapping. Diagnostic tools including [SuSiE-RSS][susierss_diagnostic], and [DENTIST][DENTIST], have been developed to check possible LD mismatch. We have provided such analysis in cTWAS. However, it is time consuming to run the LD mismatch diagnosis for all the regions across the genome, so we will perform the diagnosis and adjustment only for selected regions with high PIP signals in the post-processing step. Please see the tutorial "Post-processing cTWAS results" for details.

Below we describe several functions to harmonize the input datasets. 

### Harmonizing GWAS z-scores and the reference data

The `preprocess_z_snp()` function harmonizes GWAS z-scores and the reference data based on the included allele information.

```{r preprocess_z_snp, eval=FALSE}
z_snp <- preprocess_z_snp(z_snp, snp_map, 
                          drop_multiallelic = TRUE, 
                          drop_strand_ambig = TRUE)
```


By default, it will drop multiallelic variants and strand ambiguous variants. You can change these by setting the options `drop_multiallelic` and `drop_strand_ambig`.

### Harmonizing prediction models and the reference data

We use the `preprocess_weight()` function to harmonize the PredictDB/FUSION prediction models and LD reference. 

```{r preprocess_weights, eval=FALSE}
weights <- preprocess_weights(weight_file,
                              region_info,
                              z_snp$id,
                              snp_map,                                  
                              weight_format = "PredictDB",
                              drop_strand_ambig = TRUE,
                              scale_predictdb_weights = TRUE,
                              load_predictdb_LD = TRUE,
                              filter_protein_coding_genes = TRUE,
                              ncore = 6)
```


This function returns `weights`, which contains the effect sizes of the variants in the prediction models, LD of variants in the weights, the first and last eQTL positions of genes. 

In the function above, we specify the weight format in `weight_format`. We use PredictDB format by default. We set `load_predictdb_LD = TRUE` to load pre-computed covariances between variants in PredictDB weights (`.txt.gz`). PredictDB weights assume that variant genotypes are not standardized, but our implementation assumes standardized variant genotypes. Thus, we set `scale_predictdb_weights = TRUE`, to scale PredictDB weights by the variance before computing Z-scores of genes. If weights are already on the standardized scale (e.g. FUSION weights or PredictDB weights converted from eQTLs), this scaling should be turned off.

We drop strand ambiguous variants (`drop_strand_ambig = TRUE`) by default. We could limit to protein coding genes (`filter_protein_coding_genes = TRUE`). We could set `ncore` to parallelize the computation.

[reference]: https://xinhe-lab.github.io/ctwas/reference/index.html
[UKBB_LD_ref]: https://uchicago.box.com/s/jqocacd2fulskmhoqnasrknbt59x3xkn
[LDetect]: https://github.com/endrebak/ldetect
[PredictDB]: http://predictdb.org/
[FUSION_format]: http://gusevlab.org/projects/fusion/#compute-your-own-predictive-models
[S-PrediXcan]: https://www.nature.com/articles/s41467-018-03621-1
[susierss_diagnostic]: https://stephenslab.github.io/susieR/articles/susierss_diagnostic.html
[DENTIST]: https://github.com/Yves-CHEN/DENTIST/
